---
title: "hw3"
author: "Mark Vandergon"
date: "10/14/2017"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(gganimate)
library(readxl)
library(lubridate)
library(showtext)

# ggplot theme
font_add("Nexa Light", "Nexa Light.otf")
font_add("Nexa Bold", "Nexa Bold.otf")

###################
## theme

pallete<- c("#484EF4", "#1922FF",  "#0009FF",	"#B29000", 	"#FFCD00")
pallete2 <- c("#4870F4", "#3D5399",  "#00C2FF",	"#FF8440", 	"#CC3912")
scale_fill_mark <- function(){
  structure(list(
    scale_fill_manual(values=pallete)
  ))
}

scale_color_discrete_mark <- function(){
  structure(list(
    scale_color_manual(values=pallete)
  ))
}

scale_color_continuous_mark <- function(){
  structure(list(
    scale_color_gradientn(colours = pallete)
  ))
}
##################################

theme_mark <- function(base_size=12, font=NA){
  
  txt <- element_text(size = base_size, colour = "black", family="Nexa Light")
  bold_txt <- element_text(size = base_size+2, colour = "black", family="Nexa Bold")
  
  theme_minimal(base_size = base_size, base_family = font) +
    theme(
      ###### clean up!
      legend.key = element_blank(), 
      strip.background = element_blank(), 
      ########### text basics
      text = txt, 
      plot.title = bold_txt, 
      axis.title = txt, 
      axis.text = txt, 
      legend.text = txt ) +
    
    ############## axis lines
    theme(
      axis.line.y = element_line(colour = "#333333", size = 0.8, linetype = "solid"),
      axis.line.x = element_line(colour = "#333333", size = 0.8, linetype = "solid"),
      
      ### legend  top and no title!
      legend.position = "top", 
      legend.title = element_blank(),
      legend.key = element_rect(fill = "#f8f8f8", color = "#f8f8f8"),
      legend.background = element_rect( fill = "#f8f8f8",color = "#f8f8f8", size = 0.5, linetype = "solid"),
      
      ## background
      plot.background = element_rect(fill = "#f0f0f0",colour = "#f0f0f0",size = 0.5)
      )
}

####################

theme_set(theme_mark()) 
```

```{r loading_data echo=FALSE, message=FALSE}
# setwd("~/data-viz/hw3")
#### load the data

# Dataset 1
# FRED Federal Reserve Bank of St. Louis, selected metrics (FEDFUNDS, GDPC1, GDPDEF, GDPPOT, and CPILFESL)
fedfunds <- read_csv('FEDFUNDS.csv')
fedfunds$DATE <- ymd(fedfunds$DATE)
fedfunds <- filter(fedfunds, DATE > "1960-01-01" & DATE <"2017-07-01")
gdpc1 <- read_csv('GDPC1.csv')
gdpc1$DATE <- ymd(gdpc1$DATE)
gdpc1 <- filter(gdpc1, DATE > "1960-01-01" & DATE <"2017-07-01")
gdpdef.pc1 <- read_csv('GDPDEF_PC1.csv')
gdpdef.pc1$DATE <- ymd(gdpdef.pc1$DATE)
gdpdef.pc1 <- filter(gdpdef.pc1, DATE > "1960-01-01" & DATE <"2017-07-01")
gdppot <- read_csv('GDPPOT.csv')
gdppot$DATE <- ymd(gdppot$DATE)
gdppot <- filter(gdppot, DATE > "1960-01-01" & DATE <"2017-07-01")

# combine for taylor rate
taylor <- inner_join(by='DATE', gdpc1, gdpdef.pc1) %>%
          inner_join(by='DATE', gdppot)

# track against core CPI
core_cpi <- read_csv('CPILFESL.csv')
core_cpi$DATE <- ymd(core_cpi$DATE)
core_cpi <- filter(core_cpi, DATE > "1960-01-01" & DATE <"2017-07-01")

# Dataset 2
# Cavallo, Alberto, and Roberto Rigobon (2016) "The Billion Prices Project"
pp <- read_csv('pricestats_bpp_arg_usa.csv')
pp$date <- dmy(pp$date)
t <- c(1000)
pp_recent <- pp %>%
              filter(country == "USA") %>%
              top_n(t, date) 
```

## Data and Metrics
#### Effective Federal Funds Rate
The federal funds rate is the interest rate at which depository institutions trade federal funds (balances held at Federal Reserve Banks) with each other overnight. When a depository institution has surplus balances in its reserve account, it lends to other banks in need of larger balances. In simpler terms, a bank with excess cash, which is often referred to as liquidity, will lend to another bank that needs to quickly raise liquidity.

#### The Taylor Rule:
https://fred.stlouisfed.org/graph/?g=cN69
It has three main inputs:current economic output, potential output and inflation. Often a 2% steady state inflation is assummed. In future assignments, I may use realtime data to get a more granular estimate of inflation.

a. Real Potential Gross Domestic Product: GDPPOT
b. Real Gross Domestic Product: GDPC1
c. % Annual Change of Gross Domestic Product: Implicit Price Deflator (indexed): GDPDEF_PC1

The Taylor Rule formula for interest rate:
``` {r taylor_rule error=FALSE, echo=FALSE} 
taylor_rate <- function (potential_gdp, real_gdp, gdpdef_pc1, inflation=2){
  rate <- gdpdef_pc1+inflation +
          0.5*(gdpdef_pc1-inflation) +
          0.5*(real_gdp-potential_gdp)/potential_gdp
  return(rate)
}
taylor <- taylor %>% 
          rowwise() %>% 
          mutate(RATE= taylor_rate(GDPPOT,GDPC1,GDPDEF_PC1))
```

Plotting against the real rate, we can observe how well Taylor tracks with the True
``` {r rates, echo=FALSE}
rates <- ggplot() +
      geom_line(data = taylor, aes(x=DATE, y=RATE), color = "#FF3E23", size=2) +
      geom_line(data = fedfunds, aes(x=DATE, y=FEDFUNDS), color = "#333333") +
      labs(y = "Interest Rate (%)",
           x = "Date",
           title="The Taylor Rule loosely tracks with the Real Fed Funds Rate",
           subtitle="Simplified Taylor Rule interest rate against real fed funds rate (source: FRB St. Louis)")
rates
```

```{r pp, echo=FALSE}
pp <- ggplot(pp_recent, aes(x=date, y=indexCPI, group=1)) +
  geom_line(colour = "#f45f09")+
  stat_smooth(span=0.5, color="#DDDDDD") +
  labs(x="Date", 
       y="CPI",
       title="Consumer Price Index: 2013 to 2015",
       subtitle="(source:Billion Prices)")
pp
```

Core CPI (no food or oil prices) can be used to recognize periods of inflation and deflation. Significant increases in the CPI within a short time frame might indicate a period of inflation, and significant decreases in CPI within a short time frame might indicate a period of deflation.
