---
title: "Spatial Visualization"
output:
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
Sys.setenv(TZ="America/Chicago")
library(tidyverse)
library(lubridate)
library(tigris)
library(ggthemes)
library(tilegramsR)
library(leaflet)
library(leaflet.extras)
library(colormap)

#import a custom map theme
source("../custom_map_theme.R")

# Leaflet Options
getLeafletOptions <- function(minZoom, maxZoom, ...) {
  leafletOptions(
    crs = leafletCRS("L.CRS.Simple"),
    minZoom = minZoom, maxZoom = maxZoom,
    dragging = FALSE, zoomControl = FALSE,
    tap = FALSE,
    attributionControl = FALSE , ...)
}
getFactorPal <- function(f) {
  colorFactor(colormaps$viridis, f)
}
```

#### Motivation: Regional GDP Differences
There are only a few states who are exceeding their Potential GDP, or highest efficiency. [this article](https://research.stlouisfed.org/publications/economic-synopses/2013/08/13/mind-the-regional-output-gap/) from the St. Louis Federal Reserve.

``` {r loading_shapefile, echo=FALSE, message=FALSE}
# states shapefile
states <- states(cb = FALSE, resolution = "5m", year = 2017)
```

``` {r data_wrangling, echo=FALSE, message=FALSE}
# state gdp (Bureau of Economic Advisors: https://www.bea.gov/regional/downloadzip.cfm)
state_gdp <- read_csv('QGSP_ALL_R.csv') %>% 
                # filter for only all industries and actual states only (Not DC)
                filter(IndustryId == 1, GeoFIPS != '00000', GeoFIPS != '11000', GeoFIPS < 57) %>%
                # getting a column to join on
                mutate('id' = as.integer(GeoFIPS)/1000)

population <- read_csv('nst-est2016-alldata.csv') 
state_gdp <- dplyr::left_join(state_gdp, population, by = c("id"= "STATE"))

# convert quarterly data to numeric
state_gdp[, grepl( "Q" , names( state_gdp ) ) ] <- state_gdp[, grepl( "Q" , names( state_gdp ) ) ] %>%
                lapply(function(x) as.numeric(as.character(x)))

# rename columns so R doesn't throw errors
colnames(state_gdp)[grepl( "Q" , names(state_gdp))] <- paste0("D_",colnames(state_gdp)[grepl( "Q" , names( state_gdp))])

# fortify to convert into a dataframe, trim by contiguous US
top = 49.3457868 # north lat
left = -124.7844079 # west long
right = -66.9513812 # east long
bottom =  24.7433195 # south lat

state_map <- fortify(states, region="STATEFP") %>%
              # make id column numeric in order to join
              mutate(id = as.integer(id)) %>%
              filter(right >= long & long >= left, top >= lat & lat >= bottom)

states@data$STATEFP <- as.integer(states@data$STATEFP)
state_map <- dplyr::left_join(state_map, states@data, by = c("id"= "STATEFP"))

# join "thematic" data
gdpDF <- dplyr::left_join(state_map, state_gdp, by = c("id"))

#chartogram wrangle
sf_NPR1to1 <- mutate(sf_NPR1to1, FID = as.integer(FID))
sf_USGDP <- dplyr::left_join(sf_NPR1to1, state_gdp, by = c("FID"="id"))
sf_USGDP.centers <- sf_NPR1to1.centers
```
## Charts 1: ggPlot of GDP
Let's examine a recent GDP figure. In terms of magnitudes it is no surprise, mostly tied to the population of the state.

```{r gg_rgdp}

ggplot(state_gdp, aes(map_id=id,fill= D_2016Q1)) +
  geom_map(map=state_map, color='#ffffff', size=0.1) +
  expand_limits(x=state_map$long,y=state_map$lat) +
  scale_fill_colormap('GDP per Capita', discrete = F, colormap = colormaps$viridis, reverse = T)+
  theme_map_mark() +
  labs(title="California Contributes the Most to GDP",
             subtitle="Real Dollars 2016Q - 148 states",
             caption="source: Bureau of Economic Advisors")
```

## Chart 2: Per Person
But we're looking at potential GDP, so perhaps a naive way to see this is via GDP per Capita. Maybe the story will be different? 

```{r gg_rgdp_weighted}

ggplot(state_gdp, aes(map_id=id,fill= D_2016Q1/CENSUS2010POP)) +
  geom_map(map=state_map, color='#ffffff', size=0.1) +
  expand_limits(x=state_map$long,y=state_map$lat) +
  scale_fill_colormap('GDP per Person', discrete = F, colormap = colormaps$viridis, reverse = T)+
  theme_map_mark() +
  labs(title="Productivity isn't Only in Populous States",
             subtitle="Real Dollars per person 2016Q1 - 48 states",
             caption="source: Bureau of Economic Advisors")
```

## Chart 3: Interactive GDP per Person
Let's have another look at the GDP per Person in the contiguous 48, but this time, abstracting away the shape.

``` {r chartogram_rdgp}
# via tilegramsR
sf_USGDP$VALUE <- sf_USGDP$D_2016Q1/sf_USGDP$CENSUS2010POP
qpal <- colorBin("Greens", domain = sf_USGDP$VALUE, bins = 4)

labels <- sprintf(
  "<strong>%s</strong><br/>%g dollars per person",
  sf_USGDP$GeoName, sf_USGDP$VALUE 
) %>% lapply(htmltools::HTML)

leaflet(
  sf_USGDP,
  options= getLeafletOptions(-1.5, -1.5), width = "100%" )%>%
  addPolygons(
    weight=2,
    fillColor= ~qpal(VALUE),
    group = 'states',
    fillOpacity = 0.6, opacity = 1, color="#666666",
    highlightOptions = highlightOptions(weight = 4),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(title="GDP ($) per Person", pal = qpal, values = ~'2016Q1', opacity = 0.7,
    position = "bottomright")%>%
  addLabelOnlyMarkers(
    data=sf_USGDP.centers,
    label = ~as.character(state),
    labelOptions = labelOptions(
    noHide = 'T', textOnly = T,
    offset=c(-4,-10), textsize = '12px')) %>%
  setMapWidgetStyle()
```
